\begin{thebibliography}{99}

\bibitem{lei2018machinery}
Y.~Lei, B.~Yang, X.~Jiang, F.~Jia, N.~Li, and A.~K. Nandi,
``Machinery health prognostics: A systematic review from data acquisition to rul prediction,''
\textit{Mechanical Systems and Signal Processing}, vol.~104, pp.~799--834, 2018.

\bibitem{fink2020prognostics}
O.~Fink, E.~Zio, and U.~Weidmann,
``Prognostics and health management---a review on data driven approaches,''
\textit{European Journal of Operational Research}, vol.~279, no.~3, pp.~774--794, 2020.

\bibitem{mckinsey2017}
McKinsey \& Company,
``The value of predictive maintenance,''
\texttt{https://www.mckinsey.com}, 2017.

\bibitem{mobley2019}
R.~K. Mobley,
\textit{An Introduction to Predictive Maintenance}.
Elsevier, 2019.

\bibitem{nectoux2012pronostia_a}
P.~Nectoux, R.~Gouriveau, K.~Medjaher, E.~Ramasso, B.~Morello, N.~Zerhouni, and C.~Varnier,
``Pronostia: An experimental platform for bearings accelerated degradation tests,''
in \textit{2012 IEEE International Conference on Prognostics and Health Management (PHM)}, 2012.

\bibitem{saxena2008damage}
A.~Saxena, K.~Goebel, D.~Simon, and N.~Eklund,
``Damage propagation modeling for aircraft engine run-to-failure simulation,''
in \textit{Proceedings of the 1st International Conference on Prognostics and Health Management (PHM'08)}, 2008.

\bibitem{guidotti2024counterfactual}
R.~Guidotti,
``Counterfactual explanations and how to find them: literature review and benchmarking,''
\textit{Data Mining and Knowledge Discovery}, vol.~38, no.~5, pp.~2770--2824, 2024.

\bibitem{lundberg2017unified}
S.~M. Lundberg and S.-I. Lee,
``A unified approach to interpreting model predictions,''
in \textit{Advances in Neural Information Processing Systems 30}, 2017.

\bibitem{ribeiro2016trust}
M.~T. Ribeiro, S.~Singh, and C.~Guestrin,
```Why should I trust you?': Explaining the predictions of any classifier,''
in \textit{Proceedings of the 22nd ACM SIGKDD}, 2016.

\bibitem{wachter2017counterfactual}
S.~Wachter, B.~Mittelstadt, and C.~Russell,
``Counterfactual explanations without opening the black box: Automated decisions and the GDPR,''
\textit{Harvard Journal of Law \& Technology}, vol.~31, no.~2, 2017.

\bibitem{mothilal2020explaining}
R.~K. Mothilal, A.~Sharma, and C.~Tan,
``Explaining machine learning classifiers through diverse counterfactual explanations,''
in \textit{Proceedings of the 2020 ACM Conference on Fairness, Accountability, and Transparency (FAT*)},
pp.~607--617, 2020.

\bibitem{jin2025tcntransformer}
X.~Jin, Y.~Ji, S.~Li, K.~Lv, J.~Xu, H.~Jiang, and S.~Fu,
``Remaining useful life prediction for rolling bearings based on TCN--Transformer networks using vibration signals,''
\textit{Sensors}, vol.~25, no.~11, p.~3571, 2025.

\bibitem{zhou2021transformer}
Z.~T. Zhou, L.~Liu, X.~Song, et~al.,
``Remaining useful life prediction method of rolling bearing based on Transformer model,''
\textit{Journal of Beijing University of Aeronautics and Astronautics}, vol.~49, no.~2, pp.~430--443, 2021.

\bibitem{zou2023multiscale}
W.~Zou, Z.~Lu, Z.~Hu, and L.~Mao,
``Remaining useful life estimation of bearing using deep multiscale window-based transformer,''
\textit{IEEE Transactions on Instrumentation and Measurement}, vol.~72, pp.~1--11, 2023.

\bibitem{moosavi2024explainable}
S.~Moosavi, M.~Farajzadeh-Zanjani, R.~Razavi-Far, V.~Palade, and M.~Saif,
``Explainable AI in manufacturing and industrial cyber--physical systems: A survey,''
\textit{Electronics}, vol.~13, no.~17, p.~3497, 2024.

\bibitem{li2022sgcf}
P.~Li, O.~Bahri, S.~Filali Boubrahimi, and S.~M. Hamdi,
``SG-CF: Shapelet-guided counterfactual explanation for time series classification,''
in \textit{2022 IEEE International Conference on Big Data (Big Data)}, pp.~1564--1573, 2022.

\bibitem{sanderson2025dipace}
J.~Sanderson, H.~Mao, and W.~L. Woo,
``DiPACE: Diverse, plausible and actionable counterfactual explanations,''
in \textit{Proceedings of the 17th International Conference on Agents and Artificial Intelligence (ICAART 2025)},
vol.~2, pp.~543--554, 2025.

\bibitem{ding2021convolutional}
Y.~Ding and M.~Jia,
``A convolutional Transformer architecture for remaining useful life estimation,''
in \textit{2021 Global Reliability and Prognostics and Health Management (PHM-Nanjing)},
pp.~1--6, 2021.

\bibitem{duan2024deepar}
Z.~Duan et~al.,
``Intelligent prediction of rolling bearing remaining useful life based on probabilistic DeepAR-Transformer model,''
\textit{Measurement Science and Technology}, vol.~35, no.~1, p.~015107, 2024.


\bibitem{ates2021counterfactual}
E.~Ates, B.~Aksar, V.~J. Leung, and A.~K. Coskun,
``Counterfactual explanations for multivariate time series,''
in \textit{Proceedings of the 2021 International Conference on Applied Artificial Intelligence (ICAPAI)},
pp.~1--8, 2021.
DOI: 10.1109/ICAPAI49758.2021.9462056.

\bibitem{li2023abcf}
P.~Li, O.~Bahri, S.~Filali Boubrahimi, and S.~M. Hamdi,
``Attention-based counterfactual explanation for multivariate time series,''
in \textit{International Conference on Big Data Analytics and Knowledge Discovery (DaWaK)},
pp.~315--329, 2023.
DOI: 10.1007/978-3-031-39831-5\_26.

\bibitem{delaney2021instance}
E.~Delaney, D.~Greene, and M.~T. Keane,
``Instance-based counterfactual explanations for time series classification,''
in \textit{Proceedings of the 29th International Conference on Case-Based Reasoning (ICCBR)},
2021. arXiv:2009.13211.

\bibitem{bahri2022sets}
O.~Bahri, P.~Li, S.~Filali Boubrahimi, and S.~M. Hamdi,
``Shapelet-based counterfactual explanations for multivariate time series,''
in \textit{KDD Workshop on Mining and Learning from Time Series (MiLeTS)},
2022. arXiv:2208.10462.

\bibitem{heimes2008recurrent}
F.~O. Heimes,
``Recurrent neural networks for remaining useful life estimation,''
in \textit{Proceedings of the 2008 International Conference on Prognostics and Health Management (PHM)},
Denver, CO, USA, 2008, pp.~1--6.

\bibitem{zhang2011extracting}
X.~Zhang and Y.~Zhao,
``Extracting key postures with sparse auto-encoder from motion capture data,''
in \textit{Proceedings of the 22nd British Machine Vision Conference (BMVC)}, 2011.

\bibitem{vaswani2017attention}
A.~Vaswani, N.~Shazeer, N.~Parmar, J.~Uszkoreit, L.~Jones, A.~N. Gomez, {\L}.~Kaiser, and I.~Polosukhin,
``Attention is all you need,''
in \textit{Advances in Neural Information Processing Systems 30 (NeurIPS)}, 2017, pp.~5998--6008.

\bibitem{zerveas2021transformer}
G.~Zerveas, S.~Jayaraman, D.~Patel, A.~Bhamidipaty, and C.~Eickhoff,
``A transformer-based framework for multivariate time series representation learning,''
in \textit{Proceedings of the 27th ACM SIGKDD Conference on Knowledge Discovery and Data Mining},
2021, pp.~2114--2124.

\bibitem{li2018remaining}
J.~Li, X.~Li, and D.~He,
``Remaining useful life estimation in prognostics using deep convolution neural networks,''
\textit{Reliability Engineering \& System Safety}, vol.~172, pp.~1--11, 2018.

\bibitem{cpcb2020aqi}
Central Pollution Control Board (CPCB),
``Air Quality Data --- \texttt{city\_day.csv},''
Kaggle, 2020.
\texttt{https://www.kaggle.com/datasets/rohanrao/air-quality-data-in-india}.

\bibitem{bagnall2018uea}
A.~Bagnall, H.~A. Dau, J.~Lines, M.~Flynn, J.~Large, A.~Bostrom, P.~Southam, and E.~Keogh,
``The UEA multivariate time series classification archive, 2018,''
arXiv:1811.00075, 2018.

\bibitem{loshchilov2019decoupled}
I.~Loshchilov and F.~Hutter,
``Decoupled weight decay regularization,''
in \textit{Proceedings of the 7th International Conference on Learning Representations (ICLR)}, 2019.

\bibitem{abnar2020quantifying}
S.~Abnar and W.~Zuidema,
``Quantifying attention flow in transformers,''
in \textit{Proceedings of the 58th Annual Meeting of the Association for Computational Linguistics (ACL)},
2020, pp.~4190--4197.

\end{thebibliography}
